Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-05-02 thesis/data.tex

@@

Gaia provides unprecedented accuracy in proper motions and magnitudes.
Gaia data is uniquely excellent to produce low-contamination samples of
likely member stars belonging to satellites. Here, we breifly describe
J+24's membership estimation and discuss how this informs our
observational knoledge of each galaxies density profile. In general,
J+24 use a Bayesian framework incorporating proper motion (PM),
colour-magnitude diagram (CMD), and spatial information to determine the
probability that a given star belongs to the satellite. J+24 extends
\citet{MV2020a} (see also \citet{pace+li2019}, \citet{battaglia+2022},
\citet{pace+erkal+li2022}, etc.).

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  High quality astrometry (\texttt{ruwe\ \textless{}=\ 1.3}).
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance
  (typically near zero; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute RA and Dec proper motions less than
  10\(\,{\rm mas\ yr^{-1}}\) (corresponding to tangental velocities of
  \(\gtrsim 500\) km/s at distances larger than 10 kpc.).
\item
  small colour excess factor \(|C*| \leq 3\sigma_{C*}(G)\)
  (\citet{riello+2021}).
\item
  22 \textgreater{} G \textgreater{} 5\(\sigma\) brighter than TRGB
  (dereddened). In practice astrometry cuts limit 21 \textgreater{} G.
\item
  Between -0.5 and 2.5 in Bp - Rp (dereddened).
\end{itemize}

@@

J+24 calculate the probability that any star belongs to either the
satellite or the MW background with \[
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\]

where \(f_{\rm sat}\) is the fraction of candidate members in the field,
and \({\cal L}_{\rm sat}\) and \({\cal L}\)\_\{\rm bg\}\$ are the
satellite (sat) and background (bg) likelihoods respectively. Each
likelihood is the product of a spatial, proper motion, and CMD term: \[
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\]

The satellite likelihoods are specified as

\begin{itemize}
\tightlist
\item
  CMD: The CMD is the lowest metallicity isochrone from Padova
  \citep{girardi+2002} with age 12 Gyr with a colour width of 0.1 mag
  plus the Gaia colour uncertainty at each magnitude. The HB is modelled
  as a constant magnitude extending blue of the CMD. The HB magnitude is
  the mean magnitude of HB stars from most metal poor isochrone with a
  0.1 mag width plus the mean colour error. A likelihood map is
  constructed by sampling the distance modulus in addition to the CMD
  width, taking the maximum of RGB and HB likelihoods.
\item
  Spatial: A single exponential
  (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)) accounting for structural
  uncertainty (sampled over position angle, ellipticity, and half light
  radius).
\item
  Alternative spatial: For Scl and UMi, this is instead a double
  exponential
  \(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\) where
  the inner exponential remains fixed. Structural parameter
  uncertainties are not accounted for.
\item
  PM. A bivariate gaussian with variance and covariance equal to each
  star's proper motions. Each star's proper motions uncertainty are
  assumed to be the dominant uncertainty.
\end{itemize}

The background likelihoods are constructed as:

\begin{itemize}
\tightlist
\item
  CMD : Constructed as a density map using the other quality-selected
  stars outside of \(5R_h\) in the catalogue. The map is a sum of
  bivariate gaussians for each star with standard deviations based on
  the Gaia uncertainties.
\item
  PM: same as CMD except in PM space (\(\mu_{\alpha*}\) and
  \(\mu_\delta\) with respecive uncertainties and covariance.)
\item
  Spatial: a constant likelihood.
\end{itemize}

Each likelihood map is normalized over the respective 2D parameter
space. \(f_{\rm sat}\) thus is the only term controlling the total
abundance of satellite stars relative to the background.

In J+24, a MCMC simulation is ran using the above total likelihood to
solve for the following parameters:

\begin{itemize}
\tightlist
\item
  Systemic proper motions \(\mu_\alpha\), \(\mu_\delta\). Single
  component prior is same as \citet{MV2020}: a normal distribution with
  mean 0 and standard deviation 100 km/s. If 2-component spatial,
  instead is a uniform distribution spanning 5\(\sigma\) of single
  component case w/ systematic uncertainties.
\item
  \(f_{\rm sat}\) density normalization. Prior is a uniform distribution
  between 0 and 1.
\item
  Spatial component parameters \(B\) is uniform from 0-1 and
  \(R_{\rm outer}\) is uniform and greater than \(R_s\) for extended
  profiles (Scl and UMi here.)
\end{itemize}

The mode of each parameter from the MCMC are then used to calculate the
final \(P_{\rm sat}\) values we use here.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to 0 or 1, so the
choice of probability threshhold is not too significant. Additionally,
even for a probability cut of 0.2, the purity of the resulting sample
with RV measurements is very high (\textasciitilde90\%, J+24). However,
there is likely a systematic bias in using stars with RV measurements to
measure purity. Fainter stars have poorer astrometry and are less likely
to have been targeted. We find no difference in the resulting density
distributions when restricting stars to be brighter than a specific
magnitude.

@@

In figures fig.~\ref{fig:sculptor_selection},
fig.~\ref{fig:umi_selction}, we illustrate the resulting samples from
the algorithm in the tangent plane (\(\xi\), \(\eta\), \emph{does this
need defined?}), CMD, and proper motion space. For both galaxies, each
criteria plays a commensurate role in sifting out members. Proper
motions are clustered around the dwarf systemic motion, the CMD is well
defined including the horizontal branch, and stars only within a few
\(R_h\) are included. We also mark stars with consistent radial
velocities (RVs) (see below), which trace each feature albeit more
cautiously.

Figure REF also illustrats the distribution of stars selected without a
spatial criterion. We define the CMD+PM selection as stars satisfying \[
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
\] These stars are distributed similar to the fiducial (probable
members) sample but with a fainter uniform distribution across the
entire field. This illustrates the approximate background of stars which
may be confused as members. Additionally, since there is no clear
spatial structure in the CMD+PM sample, it is unlikely that there are
additional faint, tidal features detectable with Gaia. Not shown here,
we also try a variety of simpler, absolute cuts and thresholds, finding
no extended structure beyond what is detected in J+24. (\emph{Is it
worth using this to calculate an OOM upper limit on the density of tidal
tails?}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Probable members (2-component) are orange, and all field stars
(satisfying quality criteria) are in light grey. \textbf{Top:} Tangent
plane. We outline in star symbols the two stars from
\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram.
\textbf{Bottom right:} Proper motion.}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor Selection]{Similar to
fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
members outside of \(3R_h\) in black stars (from \citet{sestito+2023b}
and \citet{pace+2020} and
\citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

@@

Density profiles are an essential observational constraint for our later
simulations. To derive density profiles, we use 0.05 dex bins in log
radius (i.e.~the bins are derived from
10\^{}(minimum(logR):0.05:maximum(logR))). We remove bins at smaller
(larger) radii than the first bin to contain no stars, working outwards.
We use poisson uncertainties
(\(\delta \Sigma_i / \Sigma_i = 1/\sqrt{N_i}\) for a derived density
\(\Sigma_i\) in a bin with \(N_i\) stars). As discussed below, these
uncertainties are straightforward but are likely under-represented.

fig.~\ref{fig:scl_observed_profiles},
fig.~\ref{fig:umi_observed_profiles} show the derived density profiles
for Scl and UMi. We calculate density profiles for different selections
of stars from above: all quality stars, CMD+PM only, and the probable
member samples. In each case, all samples are the same towards the inner
regions of the satellite. Classical dwarfs dominate the stellar density
in their cores. However, the density profile \emph{all stars} plateaus
at the total background in the field at radii of 30-60 arcminutes. By
restricting stars to being most likely satellite members by CMD + PM,
the background is reduced by 1-2 dex. The CMD+PM background plateau
likely represents the density of background stars which could be
mistaken as members. Finally, the background (BG)-subtracted profile
results from subtracting the apparent background in the \emph{all}
profile.

@@

One interesting note is for UMi, there is a small ultrafaint star
cluster at a position of (-100, -15) arcmin \citet{munoz+2012}. However
this cluster does not have a bright RGB so has few stars brighter than g
mag of 22. While munoz 1 may contribute about 4 stars to the density
profile of UMi, this would only apply to the bin at XXX because of the
small half light radius. We do not subtract this out.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor for
different selection criteria. \emph{probable members} selects stars with
PSAT \textgreater{} 0.2 considering PM, CMD, and spatial, \emph{CMD+PM}
select stars more likely to be members according to CMD and PM only,
\emph{all} selects any high quality star, and \emph{BG subtracted} is
the background-subtracted density derived from high-quality
stars.}\label{fig:scl_observed_profiles}
\end{figure}

@@

In particular, in fig.~\ref{fig:umi_observed_profiles}, notice that the
PSAT method produces artifically small errorbars even when the density
is \textgreater1dex below the local background. These stars are likely
selecting stars from the statistical MW background consistent with UMi
PM / CMD, recovering the assumed density profile. As a result, the
reliability of these density profiles below the CMD+PM background may be
questionable. A more robust analysis, removing this particular density
assumption, would be required to more appropriately represent the
knowledge of the density profile as the background begins to dominate.

@@

@@

@@

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 2
arcseconds. We exclude stars not matched to Gaia for simplicity.

We combine the mean RV measurement from each study using the
inverse-variance weighted mean and standard uncertainty. \[
\bar v = \frac{1}{\sum w_i}\sum_i w_i\ v_{i} 
\] \[
\delta v = \sqrt{\frac{1}{\sum_i w_{i}^2}}
\]

where \(w_i = 1/s_i^2\), and we estimate the inter-study standard
deviation \[
s^2 = \frac{1}{\sum w_i} \sum_i w_i (v_{i} - \bar v)^2
\] We remove stars with significant velocity dispersions as measured
between or within a study: \[
P\left(\chi2(N-1) < \frac{s^2}{\delta v^2}\right) > 0.001
\] where \(s, \delta v, N\) are the weighted standard deviation,
weighted standard error, and number of observations. We apply this cut
to both within a study and between studies, which typically removes
stars with reduced \(\chi^2 \gtrsim 2\).

The combined RV likelihood is then \[
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\] where \[
{\cal L}_{\rm RV, sat} = N(\mu_{v}, \sigma_{v}^2 + (\delta v_i)^2)
\] \[
{\cal L}_{\rm RV, bg} = N(0, \sigma_{\rm halo}^2)
\]

where \(\mu_v\) and \(\sigma_v\) are the systemic velocity and
dispersion of the satellite, and \(\delta v_i\) is the individual
measurement uncertainty. Typically, the velocity dispersion will
dominate the uncertainty budget here. We assume a halo/background
velocity dispersion of a constant \(\sigma_{\rm halo} = 100\) km/s
\citep[e.g.][]{brown+2010}.

Similar to above, we retain stars with the resulting membership
probability of greater than 0.2.

Finally, we need to correct the coordinate frames for the solar motion
and on-sky size of the galaxy. The first step is to subtract out the
solar motion from each radial velocity, corresponding to a typical
gradient of \textasciitilde3 km/s across the field. The next step is to
account for the slight differences in the direction of each radial
velocity. Define the \(\hat z\) direction to point parallel to the
direction from the sun to the centre of the dwarf galaxy. Then if
\(\phi\) is the angular distance between the centre of the galaxy and
the individual star, the corrected radial velocity is then \[
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
\] where
\(v_{\rm tan, R} = d(\mu_{\alpha*}\cos\theta + \mu_\delta \sin \theta)\)
is the radial component of the proper motion with respect to the centre
of the galaxy.This correction is of the order \(v_{\rm tan}\theta\) so
induces a gradient of about \(1 km/s/degree\) for sculptor
\citep[see][]{walker+mateo+olszewski2008}. The uncertainty is then the
velocity uncertainty plus the distance uncertainties times the PM
uncertainty from above. We then use the \(v_z\) values for the following
modelling, however repeating with uncorrected, heliocentric velocities
does not significantly affect the results .

>>>>

\section{Abstract}\label{abstract}

To provide an empirical benchmark for simulations in the next chapters,
we would like to understand and determine the observed properties of
Sculptor and Ursa Minor. In this section, we first review past
measurements of each galaxy and the observed properties of each dwarf
galaxy. We describe the \citet{jensen+2024} Bayesian methodology to
select high probability member stars from \emph{Gaia}. We test the
derived density profiles for Sculptor, Ursa Minor, and Fornax, finding
that \citet{jensen+2024}'s methodology successfully reduces the
background and produces similar results to a simple
background-subtracted density model. We show that Fornax is
approximately represented by a 2-D exponential profile. Instead,
Sculptor and Ursa Minor show an excess of stars relative to an 2-D
exponential in the outer regions. A Plummer profile provides a better
fit instead. Finally, we combine various radial velocity samples with
\citet{jensen+2024}'s members to derive systemic velocities and
dispersions for Sculptor and Ursa Minor.

\section{Introduction}\label{introduction}

The determination of basic properties of astronomical objects is an
ongoing and challenging endeavour.

\subsection{Gaia}\label{gaia}

Some of the most fundamental properties of astronomical objects are
their position and velocity. Unfortunately, determining positions to
stars is nontrivial, . Additionally, while line-of-sight or radial
velocities (RVs) are easily determined from spectroscopy, the tangental
velocities, perpendicular to RV, are only measurable through proper
motions, typically requiring precise astrometry as well. \emph{Gaia}'s
mission is to produce extremely precise astrometry enabling measurements
of unprecedented accuracy and scale for proper motions, parallaxes, and
magnitudes.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.3x0.75m mirrors situated at the Sun-Earth L2 lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in XXX,
completing its mission in 2025 (but with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5degrees, Gaia is able to measure
absolute proper motions by comparing the apparent shifts of stars in
different regions of the sky. In addition to precise astrometric
information, \emph{Gaia} measures the magnitude of stars in the very
wide \emph{G} band (330-1050nm), blue and red colours using the blue and
red photometers (BP and RP), and takes low resolution BP-RP spectra and
radial velocity measurements of bright stars (magnitudes \textgreater{}
16?).

\emph{Gaia} has revolutionized many astronomical disciplines, the least
of which is local group and Milky Way science. While proper motions of a
dwarf galaxies has been measured in a case by case basis by the Hubble
Space Telescope, a full systematic determination of proper motions for
most dwarf galaxies was unavailable until \emph{Gaia}
\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
the detection of many substructures of the halo, streams,

While \emph{Gaia} has performed excellently, a few limitations and
systematic uncertainties could become problematic. Gaia does show
systematic variations in its derived proper motions and parallaxes, as
measured by background quasars \citep{lindegren+2021}. These errors
limit the ability to determine absolute proper motions to some degree,
but since we would like to measure if a star is \emph{consistent} with a
dwarf galaxy, these only factor into the uncertainty budget for orbital
motion and are included in J+24. Additionally, while \emph{Gaia} was
designed to be complete down to about a magnitude of \(G=20\),
overcrowding or dense fields can limit this. \emph{Gaia} can only
successfully separate stars at a distance of about \textbf{X} (a few
times the pixel length of \ldots; pixel width is in scanning direction).
For the galaxies Ursa Minor, Fornax, and Sculptor, the highest density
(in FOrnax), is only \textbf{xxx} which appears to not significantly
affect \emph{Gaia}'s survey completeness. We test for brightness affects
in \textbf{REF}.

For our purposes, \emph{Gaia}

\citet[hereafter J+24]{jensen+2024} utilize \emph{Gaia} data with a
Bayesian framework to determine samples of high-probability members for
many Milky Way group dwarf galaxies.

Here, we breifly describe J+24's membership estimation and discuss how
this informs our observational knoledge of each galaxies density
profile. In general, J+24 use a Bayesian framework incorporating proper
motion (PM), colour-magnitude diagram (CMD), and spatial information to
determine the probability that a given star belongs to the satellite.
J+24 extends \citet{MV2020a}; \citet{MV2020b}. \citep[See also similar
work by][]{pace+li2019, battaglia+2022, pace+erkal+li2022, qi+2022}

\subsection{Summary of properties}\label{summary-of-properties}

In tbls.~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props} we present
adopted observed properties for each galaxy.

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{\label{tbl:scl_obs_props}Observed properties of Sculptor. References are: 1. Ricardo R. Muñoz et al. (2018), 2. Tran et al. (2022), 3. McConnachie and Venn (2020b), 4 McConnachie and Venn (2020a). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RR lyrae) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.18 \pm 0.23\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $9.71\pm0.17\ {\rm km\,s^{-1}}$ & sec. \ref{sec:rv_results}\\
$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & XREF\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{\label{tbl:umi_obs_props}Observed properties of Ursa Minor. References are: (1) Ricardo R. Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020a), (4) average of Pace et al. (2020) and Spencer et al. (2018). }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RR lyrae) & 2\\
distance & $70.1 \pm 3.6$ kpc & 2\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & sec. \ref{sec:rv_results}\\
$\sigma_v$ & $8.8 \pm 0.2$ & sec. \ref{sec:rv_results}\\
$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

\subsection{Data}\label{data}

@@

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\), so astrometry
  is high quality.
\item
  3\(\sigma\) consistency of measured parallax with dwarf distance
  (typically near zero; with \citet{lindegren+2021} zero-point
  correction).
\item
  Absolute RA and Dec proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\),
  less than 10\(\,{\rm mas\ yr^{-1}}\). (Corresponds to tangental
  velocities of \(\gtrsim 500\) km/s at distances larger than 10 kpc.)
\item
  Corrected colour excess is within expectations:
  \(|C^*| \leq 3\,\sigma_{C^*}(G)\), with \(C^*\) and \(\sigma_{C^*}\)
  from \citet{riello+2021}. Removes stars with unreliable photometry.
\item
  \(G\) magnitude is between \(22 > G > G_{\rm TRGB} - 5\delta\rm {DM}\)
  (dereddened magnitudes). Removes very faint stars and stars
  significantly brighter than the tip of the red giant branch (TRGB)
  magnitude plus the distance modulus uncertainty \(\delta {\rm DM}\).
\item
  Colour is between \(-0.5 < {\rm BP - RP} <  2.5\) (dereddened).
\end{itemize}

@@

J+24 calculate the probability that any star belongs to either the
satellite or the MW background with \begin{equation}
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}
\end{equation}

where \(f_{\rm sat}\) is the fraction of candidate members in the field,
and \({\cal L}_{\rm sat}\) and \({\cal L}_{\rm bg}\) are the satellite
(sat) and background (bg) likelihoods respectively. The likelihoods are
the product of a spatial, proper motion, and CMD term: \begin{equation}
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
\end{equation}

Each likelihood map is normalized over the respective 2D parameter
space. \(f_{\rm sat}\) scales the likelihoods by the abundance of stars
in the satellite compared to the background.

J+24 consider both a one-component and a two-component spatial
likelihood. The one component model is constructed as a single
exponential profile (\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with
\(R_s\) fixed to the value in the table in \citet{MV2020a}.
Additionally, structural uncertainties (for position angle, ellipticity,
and scale radius) are sampled over to construct the final likelihood
map. The two-component model instead adds a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are solved for. Structural
properties are not accounted for. The PM likelihood is a bivariate
gaussian with variance and covariance equal to each star's proper
motions. (The stellar PM uncertainties is assumed and typically is the
dominant PM uncertainty.)

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age. The (gaussian) colour width is assumed to be 0.1 mag plus the
Gaia colour uncertainty at each magnitude. The HB is modelled as a
constant magnitude extending blue of the CMD (mean magnitude of
isochone's HB stars and a 0.1 mag width plus the mean colour error). A
likelihood map is constructed by sampling the distance modulus in
addition to the CMD width, taking the maximum of RGB and HB likelihoods.

The background likelihoods are instead empirically constructed. Stars
stars outside of 5\(R_h\) passing the quality cuts estimate the
background density in PM and CMD space. The density is a sum of
bivariate gaussians with variances based on Gaia uncertainties (and
covariance for proper motions). The spatial background likelihood is
assumed to be constant.

In J+24, a MCMC simulation is ran using the above total likelihood to
solve \(\mu_{\alpha*}\), \(\mu_\delta\), and \(f_{\rm sat}\). For the
two component models, the amplitude and scale radius of the excented
compopnet is also solved for. The proper motion single component prior
is same as \citet{MV2020a}: a normal distribution with mean 0 and
standard deviation 100 km/s. If 2-component spatial, instead is a
uniform distribution spanning 5\(\sigma\) of single component case w/
systematic uncertainties. Priors on \(f_{\rm sat}\) and \(B\) (if 2
component) are uniform from 0--1. \(R_{\rm outer}\) has a uniform prior
only restricting \(R_{\rm outer} > R_s\). The mode of each parameter
from the MCMC are then used to calculate the final \(P_{\rm sat}\)
values we use here.

We adopt a probability cut of \(P_{\rm sat} = 0.2\) as our fiducial
sample. Most stars are assigned probabilities close to either 0 or 1, so
the choice of probability threshold is only marginally significant.
Additionally, even for a probability cut of 0.2, the purity of the
resulting sample with RV measurements is very high (\textasciitilde90\%,
J+24). Note there is likely a systematic bias in using stars with RV
measurements to measure purity. Fainter stars are less likely to have
been targeted and have poorer astrometry and colour information. We find
no difference in the resulting density distributions when restricting
stars to be brighter than a specific magnitude.

@@

In Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection}, we
illustrate the resulting samples from the algorithm in the tangent plane
(\(\xi\), \(\eta\), \emph{does this need defined?}), CMD, and proper
motion space. For both galaxies, each criteria plays a commensurate role
in sifting out nonmembers. Proper motions are clustered around the dwarf
systemic motion, the CMD is well defined including the horizontal
branch, and stars only within a few \(R_h\) are included. We also mark
stars with consistent radial velocities (RVs) (see below), which trace
each feature albeit more cautiously. We additionally mark the furthest
confirmed RV members of each galaxy, illustrating the large extent of
each galaxy past \(R_h\).

Figs.~\ref{fig:sculptor_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the CMD+PM selection as stars satisfying
\begin{equation}
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
\end{equation} These stars are distributed similar to the fiducial
(probable members) sample but with a fainter uniform distribution across
the entire field. This illustrates the approximate background of stars
which may be confused as members. Additionally, since there is no clear
spatial structure in the CMD+PM sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
Gaia. Not shown here, we also try a variety of simpler, absolute cuts
and thresholds, finding no extended structure beyond what is detected in
J+24. (\emph{Is it worth using this to calculate an OOM upper limit on
the density of tidal tails?}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.pdf}}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Field stars (satesfying quality criteria) are light grey,
CMD+PM selected stars are blue, probable members (2-component) are
orange, and RV confirmed members are red. \textbf{Top:} Tangent plane
\(\xi, \eta\). We outline in star symbols the two stars from
\citet{sestito+2023a}. \textbf{Bottom left:} Colour magnitude diagram in
Gaia G versus BP - RP. \textbf{Bottom right:} Proper motion in
declination \(\mu_\delta\) vs RA \(\mu_{\alpha*}\)
(corrected).}\label{fig:sculptor_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.pdf}}
\caption[Ursa Minor Selection]{Similar to
Fig.~\ref{fig:sculptor_selection} except for Ursa Minor. We outline RV
members outside of \(3R_h\) in black stars (from \citet{sestito+2023b},
\citet{pace+2020} and \citet{spencer+2018}).}\label{fig:umi_selection}
\end{figure}

@@

Density profiles are an essential observational constraint for our later
simulations. To derive density profiles, we use 0.05 dex bins in log
radius. We remove bins at smaller (larger) radii than the first bin to
contain no stars, working outwards. We use symmetric poisson
uncertainties. As discussed below, these uncertainties are
straightforward but are likely under-representative.

Figs.~\ref{fig:scl_observed_profiles}, \ref{fig:umi_observed_profiles}
show the derived density profiles for Scl and UMi. We calculate density
profiles for different selections of stars from above: all quality
stars, CMD+PM only, and the probable member samples. In each case, all
samples are the same towards the inner regions of the satellite.
Classical dwarfs dominate the stellar density in their cores. However,
the density profile \emph{all stars} plateaus at the total background in
the field at radii of 30-60 arcminutes. By restricting stars to being
most likely satellite members by CMD + PM, the background is reduced by
1-2 dex. The CMD+PM background plateau likely represents the density of
background stars which could be mistaken as members. Finally, the
background (BG)-subtracted profile results from subtracting the apparent
background in the \emph{all} profile.

@@

Nearby to UMi, there is a small, likely unassociated, ultrafaint star
cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a relative
position of \((\xi, \eta) \approx(-42, -15)\) arcminutes, corresponding
to an elliptical radius of 36 arcminutes. However this cluster does not
have a bright RGB, so has few stars brighter than a \(G\) mag of 22.
While Muñoz 1 may contribute \(\sim\) 5 stars to the density profile of
UMi, this would possibly only affect 1-2 bins due to its compact size.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods.pdf}}
\caption[Sculptor density profiles]{The density profile of Sculptor for
different selection criteria, ploted as log surface dencity versus log
elliptical radius. Residuals are with respect to th interpolated
\emph{probable members} density. \emph{Probable members} selects stars
with PSAT \textgreater{} 0.2 considering PM, CMD, and spatial,
\emph{CMD+PM} select stars more likely to be members according to CMD
and PM only, \emph{all} selects any high quality star, and \emph{BG
subtracted} is the background-subtracted density derived from
high-quality stars. We mark the half-light radius ( vertical dashed
line) and the break radius (black arrow,
REF).}\label{fig:scl_observed_profiles}
\end{figure}

@@

In particular, in Fig.~\ref{fig:umi_observed_profiles}, notice that the
PSAT method produces small errorbars, even when the density is \(>1\)dex
below the local background. These stars are likely selecting stars from
the statistical MW background consistent with UMi PM / CMD, recovering
the assumed density profile. As a result, the reliability of these
density profiles below the CMD+PM background may be questionable. A more
robust analysis, removing this particular density assumption, would be
required to more appropriately represent the knowledge of the density
profile as the background begins to dominate.

@@

Finally, our approximation to the poisson uncertainties Additionally, a
more self consistent model would fit the density profile to the entire
field at once, eliminating possible misrepresentation of the
uncertainties.

@@

\subsection{Data selection}\label{data-selection}

@@

First, we crossmatch all catalogues to J+24 Gaia stars. If a study did
not report GaiaDR3 source ID's, we match to the nearest star within 1-3
arcseconds (see REF \citet{tab:rv_measurements}). We combine the mean RV
measurement from each study using the inverse-variance weighted mean
\(\bar v\), standard uncertainty \(\delta \bar v\), and (biased)
variance \(s^2\). We remove stars with significant velocity dispersions
as measured between observations in a study or between studies. By using
that \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
\(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
distribution with \(N-1\) measurements. This cut typically removes stars
with reduced chi-squared values
\(\tilde\chi^2  = \frac{s^2}{\nu\,\delta \bar v^2}\gtrsim 7\) (since the
number of measurements is 1-3 typically).

Next, we need to correct the coordinate frames for the solar motion and
on-sky size of the galaxy. We transform the frame into the galactic
standard of rest (GSR). The next step is to account for the slight
differences in the direction of each radial velocity. Let the \(\hat z\)
be the direction from the sun to the dwarf galaxy. Then if \(\phi\) is
the angular distance between the centre of the galaxy and the individual
star, the corrected radial velocity is then \begin{equation}
v_z = v_{\rm los, gsr}\cos\phi  - v_{\alpha}\cos\theta \sin\phi - v_\delta \sin\theta\sin\phi
\end{equation} where \(v_{\rm los, gsr}\) is the line of sight velocity
in the GSR frame, \(v_\alpha\) and \(v_\delta\) are the tangental
velocities in RA and Dec, and \(\theta\) is the position angle of the
star with respect to the centre of the dwarf. The correction from both
effects induces an apparent gradient of about \(1.3\,\kmsdeg\) for
Sculptor and less for Ursa Minor \citep[see
also][]{WMO2008, strigari2010}. We add the uncertainty in \(v_z\) from
the distance uncertainty and velocity dispersion in quadrature to the RV
uncertainties for each star. We then use the \(v_z\) values for the
following modelling, however repeating with uncorrected, heliocentric
velocities does not significantly affect the results.

The combined likelihood, including RV information, becomes
\begin{equation}
{\cal L} = {\cal L}_{\rm space} {\cal L}_{\rm CMD} {\cal L}_{\rm PM} {\cal L}_{\rm RV}
\end{equation} where we assume that the satellite and background
distributions are Gaussian. Specifically, \begin{equation}
\begin{split}
{\cal L}_{\rm RV, sat} &= f\left( \frac{v_i -\mu_{v}}{\sqrt{\sigma_{v}^2 + (\delta v_i)^2}}\right) \\
{\cal L}_{\rm RV, bg} &= f\left( v_i /  \sigma_{\rm halo} \right)
\end{split}
\end{equation} where \(f\) is the probability density of a standard
normal distribution, \(\mu_v\) and \(\sigma_v\) are the systemic
velocity and dispersion of the satellite, and \(\delta v_i\) is the
individual measurement uncertainty. Typically, the velocity dispersion
will dominate the uncertainty budget here. We assume a halo/background
velocity dispersion of a constant \(\sigma_{\rm halo} = 100\,\kms\)
\citep[e.g.][]{brown+2010}.
